A generalized definition of multidimensional item response theory parameters

### 2024-09-06 output/Main_text.Rmd
Corrects use of saxon genitive in references

Following @reckase_difficulty_1985’s procedure,
we must find in the first place
the point of maximum slope of $`r IRF_ABBR`$ in $`r LATENT_SPACE`$.
Because the slope is dependent on the direction considered,
we need to reparameterize Equation \@ref(eq:M2PL-formulation)
in polar coordinates, as in @reckase_difficulty_1985.
However, Reckase's Equation 3 implicitly assumes orthogonality.
To contemplate the general case,
we need to introduce a few linear algebra results before proceeding further.

@@

Let $`r TEST_SPACE`$ be the vector space made up by
the set of $`r DISCR_VECTOR`$ vectors,
with the operations sum and product by a scalar as in $`r REAL_N_SPACE`$.
We want @reckase_difficulty_1985's definition of
the multidimensional parameters in rectangular coordinates to be
a particular case of our more general definition;
we can thus consider an orthonormal basis $`r TEST_SPACE_STD_BASIS`$
in $`r TEST_SPACE`$ such that $`r DISCR_STD_COORDS_ALT_EQ`$
corresponds in the model to $`r TRAIT_ORTH_COORDS`$.
Since the response probability to the `r MODEL_ACRONYM`,
given by Equation \@ref(eq:M2PL-formulation), must be invariant with respect to
the change of basis expressed in Equation \@ref(eq:change-basis),
$`r INVARIANCE_EQ`$.
Therefore, we find a basis $`r TEST_SPACE_BASIS_EQ`$ in $`r TEST_SPACE`$
defined by

@@

As we did with $`r LS_ORTH_BASIS`$ before,
we can conveniently assume that the inner product in
$`r TEST_SPACE_STD_BASIS`$ is the dot product.
Then, as both $`r TEST_SPACE_STD_BASIS`$ and $`r LS_ORTH_BASIS`$ are
orthonormal in $`r REAL_N_SPACE`$,
to make @reckase_difficulty_1985's definition a particular case of ours we can,
without loss of generality  (i.e., applying an arbitrary rotation),
consider that $`r BASES_EQ`$.
As $`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
invertible, so $`r INNER_PROD_MATRIX`$ is, and $`r INNER_PROD_MATRIX_INV_EQ`$.
The norm of $`r DISCR_VECTOR`$ thus fulfills

@@

As Equation \@ref(eq:M2PL-formula-polar) shows, the inner product matrix is
involved in the expression of the model in polar coordinates.
Importantly, this shows that @reckase_difficulty_1985's Equation 3 (p. 403)
already assumes the orthonormality of the latent space, as we stated before.

@@

We are thus considering the direction cosines with respect to the orthonormal
axes, so we can just apply @reckase_difficulty_1985's results, obtaining

@@

One may consider that the formal derivations presented here are
overly complicated,
when one might simply transform the space to rectangular Cartesian coordinates
with Equations \@ref(eq:change-basis) and \@ref(eq:discr-vector-std) and
then compute the parameters in the transformed space using
@reckase_difficulty_1985's and @reckase_discriminating_1991's procedure.
After all, interpreting the axes of this space as substantive latent traits
depends on "the distinction between coordinate axes and the constructs that are
the target of the instrument"
(Reckase, M.D., personal communication, 28 April 2015).
Such a procedure will of course yield identical results,
given the invariance of the parameters to a change of basis.
However, the computations in the oblique space allow us to obtain
the direction cosines relative to the original latent dimensions,
which may give relevant information if these are actual, substantive traits.
Computing the parameters in the orthonormalized (or just orthogonalized) space,
by contrast, would require computing the cosines in
the original basis afterwards.

>>>>

Following Reckase's [-@reckase_difficulty_1985] procedure,
we must find in the first place
the point of maximum slope of $`r IRF_ABBR`$ in $`r LATENT_SPACE`$.
Because the slope is dependent on the direction considered,
we need to reparameterize Equation \@ref(eq:M2PL-formulation)
in polar coordinates, as in @reckase_difficulty_1985.
However, Reckase's Equation 3 implicitly assumes orthogonality.
To contemplate the general case,
we need to introduce a few linear algebra results before proceeding further.

@@

Let $`r TEST_SPACE`$ be the vector space made up by
the set of $`r DISCR_VECTOR`$ vectors,
with the operations sum and product by a scalar as in $`r REAL_N_SPACE`$.
We want Reckase's [-@reckase_difficulty_1985] definition of
the multidimensional parameters in rectangular coordinates to be
a particular case of our more general definition;
we can thus consider an orthonormal basis $`r TEST_SPACE_STD_BASIS`$
in $`r TEST_SPACE`$ such that $`r DISCR_STD_COORDS_ALT_EQ`$
corresponds in the model to $`r TRAIT_ORTH_COORDS`$.
Since the response probability to the `r MODEL_ACRONYM`,
given by Equation \@ref(eq:M2PL-formulation), must be invariant with respect to
the change of basis expressed in Equation \@ref(eq:change-basis),
$`r INVARIANCE_EQ`$.
Therefore, we find a basis $`r TEST_SPACE_BASIS_EQ`$ in $`r TEST_SPACE`$
defined by

@@

As we did with $`r LS_ORTH_BASIS`$ before,
we can conveniently assume that the inner product in
$`r TEST_SPACE_STD_BASIS`$ is the dot product.
Then, as both $`r TEST_SPACE_STD_BASIS`$ and $`r LS_ORTH_BASIS`$ are
orthonormal in $`r REAL_N_SPACE`$,
to make Reckase's [-@reckase_difficulty_1985] definition
a particular case of ours we can,
without loss of generality  (i.e., applying an arbitrary rotation),
consider that $`r BASES_EQ`$.
As $`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
invertible, so $`r INNER_PROD_MATRIX`$ is, and $`r INNER_PROD_MATRIX_INV_EQ`$.
The norm of $`r DISCR_VECTOR`$ thus fulfills

@@

As Equation \@ref(eq:M2PL-formula-polar) shows, the inner product matrix is
involved in the expression of the model in polar coordinates.
Importantly, this shows that
Reckase's [-@reckase_difficulty_1985] Equation 3 (p. 403)
already assumes the orthonormality of the latent space, as we stated before.

@@

We are thus considering the direction cosines with respect to the orthonormal
axes, so we can just apply Reckase's [-@reckase_difficulty_1985] results,
obtaining

@@

One may consider that the formal derivations presented here are
overly complicated,
when one might simply transform the space to rectangular Cartesian coordinates
with Equations \@ref(eq:change-basis) and \@ref(eq:discr-vector-std) and
then compute the parameters in the transformed space using
Reckase's [-@reckase_difficulty_1985] and
Reckase and McKinley's [-@reckase_discriminating_1991] procedure.
After all, interpreting the axes of this space as substantive latent traits
depends on "the distinction between coordinate axes and the constructs that are
the target of the instrument"
(Reckase, M.D., personal communication, 28 April 2015).
Such a procedure will of course yield identical results,
given the invariance of the parameters to a change of basis.
However, the computations in the oblique space allow us to obtain
the direction cosines relative to the original latent dimensions,
which may give relevant information if these are actual, substantive traits.
Computing the parameters in the orthonormalized (or just orthogonalized) space,
by contrast, would require computing the cosines in
the original basis afterwards.
